The Influence of Radial Drift and Gas Accretion on the C/O Ratio in Protoplanetary Disks

### 2016-04-14 Dissertate-Harvard-LaTeX/chapters/chapter1.tex
upd

Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.  This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool. \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity. These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase. The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.

@@

In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.  These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.

@@

\subsection{Hydrodynamic Effects}\label{sec:hydro}
The neglect of hydrodynamical effects in our model is best discussed in terms of the thermal mass, $M_{\rm th}$, and the length scales introduced in \S\ref{sec:scales}.  In the low mass regime, $M\pla < M_{\rm th}/ \sqrt{3}$, where $\RB < \RH$, we assume that hydrostatic balance holds out to the outer boundary at $\RH$.   In this low mass regime, \citet{Orm13} calculated the 2D (radial and azimuthal) flow patterns driven by stellar tides and disk headwinds.     On scales $\gtrsim \RB$ the flows no longer circulate the planet: they belong to the disk.  While the density structure still appears roughly spherical and hydrostatic, these flows could affect the planet's cooling.  We expect such effects to be weak, as heat losses at greater depths dominate planetary cooling, but more study is needed, especially in 3D.

@@

At higher masses, non-hydrostatic effects become more severe.  At $M\pla \gtrsim M_{\rm th}$ planets can open significant gaps \citep{zhu13}.  At yet higher masses accretion instabilities could occur \citep{AylBat12}.  However, in this high mass regime, the spherically symmetric approximation has already broken down.

@@

The importance of envelope opacity is an established factor in core accretion calculations (e.g., \citealt{stevenson82}, \citealt{ikoma00}, R06), and several studies explore the influence of opacity on the timescales for giant planet formation in detail (e.g., \citealt{HubBod05}). Treating dust (and total) opacities as a power law in temperature is a simplification.  Opacities drop by order unity when ice grains sublimate for $T \gtrsim 150$ K and they drop by orders of magnitude when silicate grains evaporate above $T \gtrsim 1500$ K \citep{semenov03, FerAle05}.  Grain growth and composition also affect opacities. Our scenario of no ongoing accretion of solids may result in a grain-free atmosphere, which could substantially reduce the critical core mass ($\sim$$1M_{\oplus}$ in the case of Jupiter, \citealt{hori10}). Paper II explores more realistic opacity laws.

For this work, we justify a simplified dust opacity by the cool temperatures of both the outer disk and our nearly isothermal radiative zones (see \S\ref{sec:profiles}).  While  convective interiors get significantly hotter than 1500 K, opacity does not affect the structure of adiabatic convecting regions.   A possible caveat is the existence of  radiative zones sandwiched inside the convection interior.  Such ``radiative windows" arise if the opacity drop from ice and metal grain sublimation occurs at sufficiently low pressures.  Our two-layer model ignores this possibility, but radiative windows are known to exist in hot Jupiter models \citep{burrows97, ab06} and have been seen in some core accretion models (Lissauer, personal communication).  The role of radiative windows in core accretion models remains to be explored in detail.

>>>>

Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.  This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool. \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity. These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase. The core accretion model of \citet{hubickyj05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.

@@

In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{rafikov11}.  These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance claimed by \cite{rafikov11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.

@@

\subsection{Hydrodynamic Effects}\label{sec:hydro}
The neglect of hydrodynamical effects in our model is best discussed in terms of the thermal mass, $M_{\rm th}$, and the length scales introduced in \S\ref{sec:scales}.  In the low mass regime, $M\pla < M_{\rm th}/ \sqrt{3}$, where $\RB < \RH$, we assume that hydrostatic balance holds out to the outer boundary at $\RH$.   In this low mass regime, \citet{ormel13} calculated the 2D (radial and azimuthal) flow patterns driven by stellar tides and disk headwinds.     On scales $\gtrsim \RB$ the flows no longer circulate the planet: they belong to the disk.  While the density structure still appears roughly spherical and hydrostatic, these flows could affect the planet's cooling.  We expect such effects to be weak, as heat losses at greater depths dominate planetary cooling, but more study is needed, especially in 3D.

@@

At higher masses, non-hydrostatic effects become more severe.  At $M\pla \gtrsim M_{\rm th}$ planets can open significant gaps \citep{zhu13}.  At yet higher masses accretion instabilities could occur \citep{ayliffe12}.  However, in this high mass regime, the spherically symmetric approximation has already broken down.

@@

The importance of envelope opacity is an established factor in core accretion calculations (e.g., \citealt{stevenson82}, \citealt{ikoma00}, R06), and several studies explore the influence of opacity on the timescales for giant planet formation in detail (e.g., \citealt{hubickyj05}). Treating dust (and total) opacities as a power law in temperature is a simplification.  Opacities drop by order unity when ice grains sublimate for $T \gtrsim 150$ K and they drop by orders of magnitude when silicate grains evaporate above $T \gtrsim 1500$ K \citep{semenov03, ferguson05}.  Grain growth and composition also affect opacities. Our scenario of no ongoing accretion of solids may result in a grain-free atmosphere, which could substantially reduce the critical core mass ($\sim$$1M_{\oplus}$ in the case of Jupiter, \citealt{hori10}). Paper II explores more realistic opacity laws.

For this work, we justify a simplified dust opacity by the cool temperatures of both the outer disk and our nearly isothermal radiative zones (see \S\ref{sec:profiles}).  While  convective interiors get significantly hotter than 1500 K, opacity does not affect the structure of adiabatic convecting regions.   A possible caveat is the existence of  radiative zones sandwiched inside the convection interior.  Such ``radiative windows" arise if the opacity drop from ice and metal grain sublimation occurs at sufficiently low pressures.  Our two-layer model ignores this possibility, but radiative windows are known to exist in hot Jupiter models \citep{burrows97, arras06} and have been seen in some core accretion models (Lissauer, personal communication).  The role of radiative windows in core accretion models remains to be explored in detail.
